##################################################
## Description: Fit Bayesian model for "comparing to a benchmark" and "testing for non-inferiority"
## Author: Doug Johnson, Torben Fischer
##################################################

## Step 0: Set-up

```{r}
rm(list=ls()) # clear current environment

# Working directory
main_path="C:/Users/tofis/Dropbox (IDinsight)/IE Design for DFEs/Code" # working directory, specify your path here. 
library(fs) # use of path()
outpath = path(main_path,"03_Output") # define output path folder,make sure sub-folder exists


# Load packages 

library(rstan)
# Stan Options
rstan_options(auto_write = TRUE)             # avoid recompilation of models
options(mc.cores = parallel::detectCores()-1)  # parallelize across all CPUs
Sys.setenv(LOCAL_CPPFLAGS = '-march=native') # improve execution time

library(rstudioapi) # to fit rstan
library(ggplot2)
library("ggpubr")

library("tibble")
library("tidyverse")    ## load the core tidyverse packages, incl. dplyr


```

## Step 1 - Data Generating Process 
## ############################################################
We first generate fake data which we will use to fit our model. 

```{r}

# Specify Data Generating Process
fake_data = function(input){  
  SEED = input[1]
  a = input[2] # treatment effect 1 
  b = input[3] # treatment effect 2 
  SIGMA = input[4]  # sd of error term/outcome, assume N(0,sigma)
  n = input[5] # sample size 
  
  set.seed(SEED)
  eps = rnorm(n,mean=0,sd=SIGMA)
  set.seed(SEED)
  T2 = rbinom(n,1,0.5) # random draw of treatment status 1 (binary)
  y = a + b*T2 + eps   # outcome data, assume simple linear and separable 
  fak_dat = data.frame(y,T2) 
  return(fak_dat)
}  

input = c(20200720,-1.2,0.3,1,150)
sim_data = fake_data(input)
sum_sim_data = sim_data %>% group_by(T2) %>% dplyr::summarize(mean_t = mean(y,na.rm=T),sd_t2 = sd(y,na.rm=T))

stan_data = list(N=input[5],y=sim_data$y,t2=sim_data$T2)

```

## Step 2 - Specify model in Stan
## ####################################
We next specify our model in Stan and save this as an R string. The Stan code below encodes the following likelihood and priors.

$$ y_i \sim N(\alpha+\theta ti,\sigma_y^2=1) $$

# Create string with details of Bayesian model in Stan
```{r}
stan_model_string = "
data {
  int<lower=0> N;  // number of observations
  vector[N] y;       // outcome variable
  vector[N] t2;     // treatment indicator
}
parameters {
  real alpha;
  real theta;
}
model {
  // priors for alpha and beta. note that the second parameter specifies the std dev not the variance (as in BUGS)
  alpha ~ normal(-1.7,2); 
  theta ~ normal(0.1,0.1); 
  y ~ normal(alpha + theta*t2,1);
} 
"
```

## Step 3 - Fit the model in Stan
```{r}
fit <- stan(model_code = stan_model_string, data = stan_data, iter = 10000, chains = 4)

# print the output from the stan function
print(fit)

# extract the posteriordraws from the fit object
extracted_values<-rstan::extract(fit, permuted = TRUE)
theta <- extracted_values$theta
```

## Step 4 - Examine output and test for convergence
Before using the Stan output, it is useful to check that Stan "converged" -- i.e. that the algorithm was successful in approximating the posterior. One useful way to do that is to make sure that the $$\hat R$$ values reported for each parameter after a call to print(fit) are less than 1.1 and, ideally, right around 1.

(COULD INSERT MORE CHECKS HERE)

## Step 5 - Perform inference using Stan output
We estimate the probability that beta is greater than .2 by calculating the % of draws for beta greater than .2

```{r}
prob_better <- sum(theta >.2)/length(theta)
print("Probability from Stan model of theta > .2")
print(prob_better)

prob_pos <- sum(theta >0)/length(theta)
print("Probability from Stan model of theta > 0")
print(prob_pos)

# kernel density of posterior for theta

p <- ggplot(data.frame(theta), aes(x=theta)) + geom_density() + geom_vline(aes(xintercept=c(0.2)))
p
ggsave(file.path(outpath, "example2_benchmark_theta_post.png"), p)
```